---
title: "Factor variables & Summary statistics in R"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes: \usepackage{setspace}\linespread{1.5}
---

```{r setup, include=FALSE}
library(knitr)
library(formatR)
library(haven)
library(tidyverse)

options(width = 80)
opts_chunk$set(echo = TRUE, out.width = 60, tidy = TRUE, tidy.opts=list(width.cutoff=80))
```

## Part 1: Factor variables

### What are factor variables and why should you use them?

Factor variables are a special format in R for categorical variables. Any data or information in categories such as gender, race or educational level can be stored in a factor variable. Factor variables can either be numeric or a string. They have two main advantages:

1.  They are easier to plot
2.  Statistical analysis: R recognizes the factor variables as categorical and uses them correctly

### Example: Voter data

Let's say we had a dataset of voters, which contains common voting survey information such as vote choice, income, education, etc. Many of these are categorical variables, so we want to make sure we create factor variables for them in R.

Let's start with vote choice. For simplicity, the 30 people in our sample voted for one of two parties.

```{r, echo = TRUE}
#create random numbers for our fictional sample
set.seed(1235) 

#randomly create a vector with 30 observations of the numbers 0 & 1:
choice <- sample(0:1, 30, replace = TRUE) 
```

Let's have a look at our data:

```{r, echo = TRUE}
choice
typeof(choice) 
```

We can see that the variable type here is numeric, so R thinks the variable contains integers 1 or 0. Obviously, we want to change them to be categories.

The below code turns our voting choice data into a categorical factor variable. We also can assign labels to the data, so that they aren't just binary numbers. As an example for our category labels, we are using the two main political parties in the UK.

```{r, echo = TRUE}
#Assign factored variable to a new name to preserve the original data
choice.f <- factor(choice, labels = c("Tory","Labour")) 
#let's have a look
choice.f 
```

\

\hrule

### Ordered factor variables

A very useful feature of factor variables in R is that they can be assigned an ordering. Of course this does not really make sense for party choice, so let's create another variable from our imaginary dataset. For example, the survey might have asked people to rank how satisfied they are with the current government's performance.

To get started, try to create a random variable vector with four categories and 30 observations. You could name it *approve* (for approval rating). Also create a factor variable from the *approve* vector and label the four categories "low", "medium", "high" and "very_high".

```{r, echo = TRUE}
## Exercise


```

R has already assigned so-called *levels* to the vector, which correspond to the labels we provided. To provide R with a specific order for the levels we can use `ordered()`.

```{r, echo = TRUE}
approve.o <- ordered(approve.f)
approve.o
```

This is again useful for certain plots and for ordered statistical models such as ordinal logistic regression.\

\hrule

### Amending factor levels

Sometimes we might realize that it makes sense to change or amend the levels. Obviously you can just reassign the variable names, but it is also easy to change. Let's use our already ordered variable:

```{r, echo = TRUE}
#first we add a level to the vector
approve.n <- factor(approve.o, levels = c(levels(approve.o), "sky_high"))

#then we change one of the data points to the new level
approve.n[6] <- "sky_high"
```

Let's drop the category again:

```{r, echo = TRUE}
approve.n <- approve.n[!approve.n == "sky_high"]
approve.n <- factor(approve.n)

```

Beware! The values that were in the sky high level have now been dropped entirely from our data. You will probably rarely want this to happen, so be careful when dropping levels. 

Finally, we might want to merge two categories. For example, it might make sense to merge the "high" and "very_high" categories:

```{r, echo = TRUE}
levels(approve.o) <- c("low", "medium","high","high")
approve.o
```

### Pitfalls

One very common issue with factor variables arises when they are conversed back to a numerical variable. Lets create an age variable which can be used as categorical or continuous. 

```{r, echo = TRUE}
age <- sample(18:99, 30, replace = TRUE) 
age.f <- factor(age)
age.reverse <- as.numeric(age.f)
```

You may already be able to spot the issue looking at the first couple of values of the two vectors in the environment tab. They are not the same at all! R has converted the factors not back to their original value, but to the numerical value of their level among the factors. So the first value in the factor vector is 6, which means that the underlying numerical value in the original data was the sixth age level.  

But not to worry, the issue is easily avoided by changing to character first:

```{r, echo = TRUE}
age.reverse2 <- as.numeric(as.character(age.f))
```

Another very common problem with factors arises from loading data into R from common file formats for data tables such as csv or txt. Prior to version 4.0 R, would load any text variables from these as factors. However now, this must be explicitly done using a command. For instance, for the `read.csv()` function, this command is `stringsAsFactors = TRUE`.


\newpage

## Part 2: Summary Statistics in R

We often need relatively simple summary statistics of our data, such as mean, median, standard deviation, etc.

To get started with this, we first need a dataset to summarize. So far, we have used slightly non-sensical random data, so let's take this up a notch. Below is a dataset of four variables from the British Election Survey, which was conducted following the UK General Election of 2019. The data has been simplified to include only those who voted Labour, Tory or Liberal Democrats.

```{r, echo = TRUE}
brits19_orig <- read.csv("brits19.csv",header = TRUE)
head(brits19_orig)
```

\

\hrule

### Exercise

Try to have a go at the following tasks:

1.  Replace the values -999, -2 and -1 with NA
2.  Convert the variables into factors where appropriate and provide labels (s. below).
3.  Provide an ordering for the factor variables where appropriate.

Below are the values for the numbers in the data:

-   vote_choice: 1 = Labour, 2 = Tory, 3 = Liberal Democrats

-   vote_mode: 1 = In person, 2 = By post, 3 = By proxy

-   income:

    -   1 = Under £5,200
    -   2 = £5,200 - £15,599
    -   3 = £15,600 - £25,999
    -   4 = £26,000 - £36,399
    -   5 = £36,400 - £46,799
    -   6 = £46,800 - £74,999
    -   7 = £75,000 - £149,999
    -   8 = £150,000 or more
    
Before you start: it is good practice to not overwrite the original data you have loaded into R. Re-assign the `brits19_orig` to simply `brits19`, which we will use going forward. 

```{r, echo = TRUE}
brits19 <- brits19_orig

## Exercise

# 1. Replace the values -999, -2 and -1 with NA


#2. Convert the variables into factors where appropriate and provide labels (s. below). 


#3. Provide an ordering for the factor variables where appropriate.


```

\

\hrule

### Simple summary statistics

Now that we have our data set up, we can begin to compute some simple summary statistics and gain an overview. There are many different ways and functions to do this in R, some of which we will briefly cover here. Frankly, the easiest thing is often to just look at an object in R Studio by clicking on it in the environment tab. 

First, let's have a look at how the data frame is structured:

```{r, echo = TRUE}
brits19 %>% str()
```

This tells us how many observations we have, as well as the number of variables and what type these are.

We can pull up the first 10 rows using`head()`:

```{r, echo = TRUE}
brits19 %>% head(10)
```

Sometimes, we will have a big data set with many different variables and want to know what these are, which we can find out using `names()`. 

```{r, echo = TRUE}
brits19 %>% names()
```

Another very useful function is `summary()`, which gives us an overview of various descriptive statistics for the entire data frame:

```{r, echo = TRUE}
brits19 %>% summary()
```

Note how R automatically computes different statistics depending on the type of variable. For the factor variables, we get a contingency table for each variable, while for the continuous variable age, we get the mean, the median as well as the minimum and maximum values.

We can also get contingency tables for each variable individually using `table()`.

```{r, echo = TRUE}
table(brits19$income)
```

If we wanted to know the percentages rather than the absolute numbers we can use `table()` inside `prop.table()`.

```{r, echo = TRUE}
prop.table(table(brits19$income))
```

We will return to tables later. Let's first run through how to compute individual statistics of interest. Because this makes less sense for our categorical variables, we will use the continuous age variable.

**Median**

```{r, echo = TRUE}
median(brits19$age, na.rm = TRUE)
```

**Mean**

```{r, echo = TRUE}
mean(brits19$age, na.rm = TRUE)
```

**Quantiles**

```{r, echo = TRUE}
quantile(brits19$age, na.rm = TRUE, 0.25) #first quartile
quantile(brits19$age, na.rm = TRUE, 0.75) #third quartile
quantile(brits19$age, na.rm = TRUE, 0.99) #99th percentile
```

**Minimum and maximum**

```{r, echo = TRUE}
min(brits19$age, na.rm = TRUE)
max(brits19$age, na.rm = TRUE)
```

**Range**

```{r, echo = TRUE}
range(brits19$age, na.rm = TRUE)
range(brits19$age, na.rm = TRUE)[1]
range(brits19$age, na.rm = TRUE)[2]
```

The range function does not actually provide us with the range, but instead gives us the two values used to compute the range, which are min and max. The result is an object that indexes the min and max. Let's get the actual range of the values, which is the difference of the maximum and minimum values:

```{r, echo = TRUE}
range(brits19$age, na.rm = TRUE)[2] - range(brits19$age, na.rm = TRUE)[1] #one way to do it
max(brits19$age, na.rm = TRUE) - min(brits19$age, na.rm = TRUE) #another way to do it
```

**Interquartile range**

```{r, echo = TRUE}
IQR(brits19$age, na.rm = TRUE)
```

**Standard deviation**

```{r, echo = TRUE}
sd(brits19$age, na.rm = TRUE)
```

**Variance**

```{r, echo = TRUE}
var(brits19$age, na.rm = TRUE)
```

**Mode**\
Another useful function is `table()`, which provides a contingency table. This can be used to compute the mode, for which there is no function in standard R.

```{r, echo = TRUE}
sort(table(brits19$age), decreasing = TRUE)[1]
```

This shows that the most frequent age in the data is 54, with 60 survey respondents being 54 years old.

\

\hrule

### Joint & conditional statistics

We are often interested in some joint statistics, like for example the voting behavior at different levels of income. This can easily be found by using `table()` or `prop.table()` and passing two variable names as arguments to the function:

```{r, echo = TRUE}
table(brits19$income, brits19$vote_choice)

round(prop.table(table(brits19$income, brits19$vote_choice)), 2) 
#note: this is rounded to the second digit 
```

Now let's say we wanted to see the voting shares of the three parties among older voters compared to all voters. It can sometimes be very laborious to perform this kind of selection or filtering in base R, so we will use use `filter()` and `select()` from the tidyverse (or more specifically: dplyr).

```{r, echo = TRUE}
over65 <- brits19 %>% 
  #we assign this to a new object so that we can use it again below
  filter(age > 65) %>% #we filter for age
  select(vote_choice) %>% #selecting the variable of interest
  table() %>% #table
  prop.table() #proportional values for vote share

all_ages <- brits19 %>% 
  #we assign this to a new object so that we can use it again below
  select(vote_choice) %>% #selecting the variable of interest
  table() %>% #table
  prop.table() #proportional values for vote share

rbind(all_ages, over65) #row-binding to see both results alongside each other


```

Note how we assigned the tables to a new variable name. This is so that we can use them again later and R knows what we mean when we refer to them. In this case, we created two tables, stored them in a newly named object, and then reused these objects inside the `rbind()` function, which allowed us to merge the two tables and see the results side by side.

Perhaps not surprisingly, the share of conservative votes (the Tory vote) is much higher among those over 65.

Tidyverse also has the `summarize()` or `summarise()` function, which allows for a more flexible computation of various summary statistics. For example, we might want to quickly access the range, mean and standard deviation of the age of Labour voters:

```{r, echo = TRUE}
brits19 %>% 
  filter(vote_choice == "Labour") %>% 
  summarize(range = max(age,na.rm = TRUE) - min(age,na.rm = TRUE), 
            mean = mean(age,na.rm = TRUE), 
            sd = sd(age,na.rm = TRUE))

```

With `mutate()` and `summarize()`, you can even compute new variables on the go.

`mutate()` is an extremely useful function from the `dplyr` package, which allows the modification of variables in a dataset. At the most basic level it lets us define a new variable like so: `mutate(new_var = )` and then requires input as to how this new variable is computed. This can also be used to overwrite old variables (although that is probably not a good idea). For example, you could create a new variable that contains the years for each voter that they have been eligible to vote: 

```{r, echo = TRUE}
brits19 %>% 
  head(10) %>% 
  mutate(new_var = age - 18) 
```

`summarize()` works similarly, but creates a data frame based on the grouping we provided (or for the entire original data frame if we don't provide one). 

Let's try this now. 

```{r, echo = TRUE}
brits19 %>% 
  filter(vote_choice == "Labour") %>% 
  mutate(first.time = (age == 18)) %>% 
  summarize(range = max(age,na.rm = TRUE) - min(age,na.rm = TRUE), 
            mean = mean(age,na.rm = TRUE), 
            sd = sd(age,na.rm = TRUE), 
            first.timers = sum(first.time, na.rm = TRUE))

```

Now we have learnt that among those that voted Labour in the survey, nine people were able to vote for the very first time! How does this compare to other parties? Try and have a go at it and see how it compares to the Tories and Liberal Democrats. Can you create a variable that gives us a proportion rather than the absolute number of first timers? (Hint: You can use `n()` to find the total number of observations.)

```{r, echo = TRUE}
## Exercise

```

Finally, can you try to find the vote share of the three parties among first time voters?

```{r, echo = TRUE}
## Exercise

```

Note that none of these operations change our original data, because we didn't assign any of it to our dataset:

```{r, echo = TRUE}
brits19 %>% 
  head()
```

\

\hrule

### Tables

You should never enjoy your statistics just by yourself. If you want to share them, it makes sense to make a nice table for them!

First, let's save the unformatted table of summary statistics and first-timer vote share.

```{r, echo = TRUE}
age_tbl <- brits19 %>% 
  group_by(vote_choice) %>% 
  #rather than using filter() and printing each, we can group by vote choice
  mutate(first.time = (age == 18)) %>% 
  summarize(range = max(age,na.rm = TRUE) - min(age,na.rm = TRUE), 
            mean = round(mean(age,na.rm = TRUE),3), 
            sd = round(sd(age,na.rm = TRUE),3), 
            first.timers = round(sum(first.time, na.rm = TRUE)/n(),3))
```

One good package for formatting our table is `kable()`. We can simply pass the table to the function:

```{r, echo = TRUE}
age_tbl %>% 
  kable() 
```

In markdown, this does not look particularly pretty. However, once we hit the `knit` button, it is formatted and produces a nice, neat-looking table.  

`kable()` also allows us to customize the output table. For example, we might want to change the number of digits that are displayed. The third digit after the comma seems unnecessary. 

```{r, echo = TRUE}
age_tbl %>% 
  kable(digits = 2)
```

Unfortunately, this cuts off the first-timer values, which are much smaller! We can pass a vector to the function to set the number of digits for each column individually: 

```{r, echo = TRUE}
age_tbl %>% 
  kable(digits = c(0,0,2,2,3)) 
#for the other variables where it doesn't apply, just pass 0s!
```

Let's modify the variable names because they are not particularly pretty the way we named them in R. Let's use the first ten rows of our original dataframe to illustrate this:

```{r, echo = TRUE}
brits19 %>% 
  head(10) %>% 
  kable(col.names = c("Vote choice", "Mode of voting", "Income level", "Age"))
```

We can also change the alignment of the columns within the function. For this, we can use a vector with the letters l, r and c for left, right and center. The letters are passed to the function as a character vector and must correspond to the columns. You can have different alignments for each column: 

```{r, echo = TRUE}
brits19 %>% 
  head(10) %>% 
  kable(col.names = c("Vote choice", "Mode of voting", "Income level", "Age"), align = "lccr")
#lccr stands for left center center right
```

To ourselves, the table can seem clear and easy to understand, but it is often useful to include a caption to make sure the reader knows what it contains as well. 

```{r, echo = TRUE}
brits19 %>% 
  head(10) %>% 
  kable(col.names = c("Vote choice", "Mode of voting", "Income level", "Age"), align = "lccr", caption = "British Election Study 2019")
```


There are a number of other packages for table formatting, such as **formattable** and **gt**. 

One very noteworthy package for publication-ready regression output tables is **stargazer**. Many people have different opinions about this package and it is sometime not exactly easy to use. But nonetheless, it produces very neat, classic regression tables which, as we all know, are the bread and butter of many results sections of empirical quantitative papers in PoliSci and beyond.

[Just to mention it: there is also the **modelsummary** package, which is easier to use but doesn't have the classic journal-style output.]

Consider a multinomial logistic regression model using our data. How this model works or what it is doesn't matter here (it basically predicts the likelihood of a voter having voted for one party rather than another). Stargazer gives us a neat table of the coefficients using the below code: 

```{r results='asis', echo = TRUE}
library(stargazer)
library(nnet)

model <- multinom(vote_choice ~ age + income + vote_mode, data = brits19) 

stargazer(model, coef = list(exp(coef(model))), covariate.labels = c("Intercept","Age","Income (medium)", "Income (high)", "Income (very high)","Mode: Post","Mode: Proxy"))
#note: for logistic models, we exponentiate the coefficients to have a more interpretable table (unless you can instantly exponentiate in your head!)

```

The below is a more vanilla example of a simple linear regression using the method of ordinary least squares. It doesn't make a whole lot of theoretical sense (we are predicting age using the other variables in the data set). Note however how stargazer can format the different models and create neat tables. 

```{r results='asis', echo = TRUE}
library(stargazer)
library(nnet)

model2 <- lm(age ~ income + vote_mode + vote_choice, data = brits19) 

stargazer(model2, covariate.labels = c("Income (medium)", "Income (high)", "Income (very high)","Mode: Post","Mode: Proxy", "Tory", "LibDem", "Constant"))#, coef = list(exp(coef(model2))))

```